GeoSPARQL 1.1: an almost decadal update to the most important geospatial LOD standard

### 2021-04-17 manuscript.tex
Corrected minor remarks of the reviewers

@@

\begin{itemize}
    \item[$\ast$] new geometry serializations
    \begin{itemize}
        \item[$-$] GeoJSON, KML etc. Newer, popular formats
    \end{itemize} 
    \item[$\ast$] new ontology classes to cater for more nuanced spatial information
    \item[$\ast$] more spatial functions
    \begin{itemize}
        \item[$-$] implementing functions well-known in non \texttt{Semantic Web} spatial systems
    \end{itemize} 
    \item[$\ast$] scalar spatial properties 
    \begin{itemize}
        \item[$-$] area, volume etc. alongside geometries
    \end{itemize} 
    \item[$\ast$] better handling of Spatial (Coordinate) Reference Systems (SRS)
    \begin{itemize}
        \item[$-$] potentially allowing for automated coordinate serialization conversions
    \end{itemize} 
    \item[$\ast$] Internet protocol-based selection of different geometries for features
\end{itemize}

@@

The reason for expecting a future, incompatible, GeoSPARQL 2.0 is that early SWG attendees thought spatio-temporal relations
and fundamental ontology elements in GeoSPARQL either could or should be remodelled, which might break the current, familiar, 
\texttt{Feature}/\texttt{Geometry} class relations. Details of these potential changes haven't been fully expounded, at the 
time of this paper, however initial SWG attendees' intuition is that a future GeoSPARQL 2.0 might generalise spatial concepts and
move away from only, or primarily, \textit{geo}spatial, or perhaps focus not just on \texttt{Feature}/\texttt{Geometry} relations
but look to generalised mechanisms for describing dimensions of features of which \textit{geometry} is one of many (and perhaps)
\textit{temporality} too).

@@

In addition to \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} defined in GeoSPARQL 1.0 
for use within SPARQL \texttt{FILTER} operations, 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
as well as \emph{geosf:BoundingCircle}, \emph{geosf:Centroid}, \emph{geosf:ConcatLines} - concatenating a set of overlapping linestrings 
that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} gives an example 
of one of these new functions in use.
% \todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions. Then I will update the namespaces here. I would find it nice to also include the functions we have discussed yesterday: transformation of SRS and in between literal formats. Should we mention it here as future work or should we make a draft pull request for that so that people have a reference? I think the latter}
\small
\begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
# returns the circle geometry bounding all the geometries 
# of Feature <x>
SELECT (geosaf:BoundingCircle(?geo) AS ?circ)
WHERE {
    <x> geo:hasGeometry/geo:asWKT ?geo .
}
\end{lstlisting}
\normalsize

@@

\section{Conclusions}\label{sec:conclusions}
A staged schedule of updates to this important \textit{Semantic Web} spatial standard has been initiated with simple and strictly backwards-compatible changes now in GeoSPARQL 1.1. Features discussed for GeoSPARQL 1.2 include the formalization of coordinate reference systems in RDF, the depiction of accurracies and level of detail and the addition of further - possibly also binary - literal types. Work on GeoSPARQL 1.2 will start later in 2021. GeoSPARQL 2.0, as yet un-specified, is likely to introduce more substantial changes to the standard. Changes proposed for GeoSPARQL 2.0 include to broaden the scope of GeoSPARQL to further kinds of spatial data. To that end, full-featured support for 3D geometries and support for coverages are discussed on the level of data representations. These proposals are related to some growing interest in the semantic web community in representing further geospatial data related to building modeling information \cite{zhang2018bimsparql} and coverage data \cite{homburg2020geosparql+}. More requirements might also be introduced once feedback has been received from the GeoSPARQL 1.1 and 1.2 releases.

>>>>

In this publication we discuss the motivations behind updating GeoSPARQL 1.0 in \Cref{sec:motivation}, the content of the planned updates for the upcoming GeoSPARQL 1.1 release in \Cref{sec:newfeatures} and finally in \Cref{sec:conclusions} provide an outlook to further feature requests which are likely to be tackled in future GeoSPARQL 1.2 and 2.0 releases.

@@

\begin{itemize}
    \item[$\ast$] new geometry serializations
    \begin{itemize}
        \item[$-$] GeoJSON, KML and other popular formats
    \end{itemize} 
    \item[$\ast$] new ontology classes to cater for more nuanced spatial information
    \item[$\ast$] more spatial functions
    \begin{itemize}
        \item[$-$] implementing functions well-known in non \texttt{Semantic Web} spatial systems
    \end{itemize} 
    \item[$\ast$] scalar spatial properties 
    \begin{itemize}
        \item[$-$] area, volume etc. alongside geometries
    \end{itemize} 
    \item[$\ast$] better handling of Spatial (Coordinate) Reference Systems (SRS)
    \begin{itemize}
        \item[$-$] potentially allowing for automated coordinate serialization conversions
    \end{itemize} 
    \item[$\ast$] Internet protocol-based selection of different geometries for features
\end{itemize}

@@

The reason for expecting a future, incompatible, GeoSPARQL 2.0 is that early SWG attendees thought spatio-temporal relations
and fundamental ontology elements in GeoSPARQL either could or should be remodelled, which might break the current, familiar, 
\texttt{Feature}/\texttt{Geometry} class relations. Details of these potential changes haven't been fully expounded, at the 
time of this paper, however initial SWG attendees' intuition is that a future GeoSPARQL 2.0 might generalise spatial concepts and
move away from only, or primarily, \textit{geo}spatial, or perhaps focus not just on \texttt{Feature}/\texttt{Geometry} relations
but look to generalised mechanisms for describing dimensions of features of which \textit{geometry} is one of many (and perhaps)
\textit{temporality} too.

@@

In addition to \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} defined in GeoSPARQL 1.0 
for use within SPARQL \texttt{FILTER} operations, 1.1 defines \href{http://www.opengis.net/def/function/geosparql/union2}{\emph{geof:union2}} and 
as well as \href{http://www.opengis.net/def/function/geosparql/boundingCircle}{\emph{geof:boundingCircle}}, \href{http://www.opengis.net/def/function/geosparql/centroid}{\emph{geof:centroid}}, \href{http://www.opengis.net/def/function/geosparql/concatLines}{\emph{geof:ConcatLines}} - concatenating a set of overlapping linestrings 
that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} gives an example 
of one of these new functions in use.
% \todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions. Then I will update the namespaces here. I would find it nice to also include the functions we have discussed yesterday: transformation of SRS and in between literal formats. Should we mention it here as future work or should we make a draft pull request for that so that people have a reference? I think the latter}
\small
\begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
# returns the circle geometry bounding all the geometries 
# of Feature <x>
SELECT (geof:BoundingCircle(?geo) AS ?circ)
WHERE {
    <x> geo:hasGeometry/geo:asWKT ?geo .
}
\end{lstlisting}
\normalsize

@@

\section{Conclusions and Outlook}\label{sec:conclusions}
A staged schedule of updates to this important \textit{Semantic Web} spatial standard has been initiated with simple and strictly backwards-compatible changes now in GeoSPARQL 1.1. Features discussed for GeoSPARQL 1.2 include the formalization of coordinate reference systems in RDF, the depiction of accurracies and level of detail and the addition of further - possibly also binary - literal types. Work on GeoSPARQL 1.2 will start later in 2021. GeoSPARQL 2.0, as yet un-specified, is likely to introduce more substantial changes to the standard. Changes proposed for GeoSPARQL 2.0 include to broaden the scope of GeoSPARQL to further kinds of spatial data. To that end, full-featured support for 3D geometries and support for coverages are discussed on the level of data representations. These proposals are related to some growing interest in the semantic web community in representing further geospatial data related to building modeling information \cite{zhang2018bimsparql} and coverage data \cite{homburg2020geosparql+}. More requirements might also be introduced once feedback has been received from the GeoSPARQL 1.1 and 1.2 releases.
